Splitting it up: the \pkg{spduration} split-population duration regression package

### 2015-12-11 manuscript.Rmd
update second section

Duration models, where the variable to be modeled describes a distribution of times until some event of interest, were originally developed in health and demographic research, where they grew naturally from life tables and survival records for medical patients. Basic formulations of such models, like the parametric exponential or Weibull regressions or semi-parametric Cox regression, implicitly assume that all subjects or units under observation, including right-censored observations, will eventually experience the event of interest. This assumption is wrong or at least disadvantageous in many substantive areas and empirical applications, where there instead is a sub-population of units or subjects that will never experience an event, and thus are effectively ``cured''.

This insight was realized as early as 1949 by @boag1949maximum and @berkson1952survival, who were researching survival rates in cancer patients, and where some fraction of patients survived because their cancer was cured, while others relapsed after apparent remission due to levels of disease below detectable thresholds. As with conventional duration models, their origin in health and medicine has shaped the terminology conventionally used with such model, e.g. split-population duration models are sometimes also referred to in such contexts as cure rate models, and the basic concepts like survival and failure rates reference the survival of humans.

Yet the intuition underlying split-population duration models has led to applications in a broad range of subject areas outside demographics and medicine. In an early and foundational application that reached beyond medicine, @schmidt1989predicting examined criminal recidivism using data on close to 10,000 prisoners from the North Carolina prison system in the late 1970's and early 1980's to identify factors that influence whether a criminal relapses at all, and if so which factors are related to the amount of time between prison stints. This work already includes a full formulation of the model with independent covariates for both the duration equation and the risk or cure equation, although only with subject-specific covariates rather than time-varying covariates with multiple data points per subject.

In a complementary set of research from public health, @douglas1994hazard use data from the US to model the age at which smokers started the habit, and @forster2001role examine the impact of tobacco taxes on smoking and quitting decisions. @deyoung2003failure, an economist, models the failure of new commercial banks in the US during the 1980's. @svolik2008authoritarian used a split-population duration framework to whether democratic regimes persist or slide back to authoritarianism, and when. Building on this effort, split-population duration models have also been used, along with other models, to produce regular predictions for five different forms of political conflict for the Integrated Crisis and Early Warning System (ICEWS) project [@ward2013learning] and to model irregular leadership changes for the Political Instability Task Force [PITF; @beger2014ensemble].

>>>>

Duration models, where baseline risk is modeled by some distribution of time, were originally developed in health and demographic research and grew naturally from life tables and survival records for medical patients. Basic formulations of such models, like the parametric exponential or Weibull regressions or semi-parametric Cox regression, implicitly assume that all subjects or units under observation, including right-censored observations, will eventually experience the event of interest. This assumption may be violated in many substantive areas and empirical applications, where a sub-population of units or individuals will never experience an event, and thus are effectively ``cured''.

The insight that populations might be split in regard to their baseline risk has been formulated as early as 1949 by @boag1949maximum and @berkson1952survival, who were researching survival rates in cancer patients. In their application some fraction of patients survived because their cancer was cured, while others relapsed after apparent remission due to levels of disease below detectable thresholds. The development of duration methods in health and medicine has also shaped the terminology conventionally used for such model. For example, split-population duration models are also referred to as cure rate models. Similarly, the basic concepts like survival and failure rates reference to the survival of humans.

Yet the intuition underlying split-population duration models has led to applications in a broad range of subject areas outside demographics and medicine. In an early and foundational application that reached beyond medicine, @schmidt1989predicting examined criminal recidivism using data on close to 10,000 prisoners from the North Carolina prison system in the late 1970's and early 1980's to identify factors that influence whether a criminal relapses at all, and if so which factors are related to the amount of time between prison stints. This work already includes a full formulation of a model with independent covariates for both the duration equation and the risk or cure equation, although only with subject-specific covariates rather than time-varying covariates for multiple data points per individual.

In a complementary set of research from public health, @douglas1994hazard use data from the United States to model the age at which individuals began to smoke, and @forster2001role examine the impact of tobacco taxes on smoking and quitting decisions. @deyoung2003failure, an economist, models the failure of new commercial banks in the US during the 1980's. @svolik2008authoritarian uses a split-population duration framework to examine whether democratic regimes persist or revert to authoritarianism, and when. Building on this effort, split-population duration models have also been used, along with other models, to produce regular predictions for five different forms of political conflict for the Integrated Crisis and Early Warning System (ICEWS) project [@ward2013learning] and to model irregular leadership changes for the Political Instability Task Force [PITF; @beger2014ensemble].
